---
title: "sph 112 measles project"
output: html_document
date: "2026-04-27"
---

```{r}
library(tidyverse)
library(readxl)
library(ggplot2)
library(janitor)
```
## Data Import
```{r}
measles_case_records <- read_xlsx("Utopia_Measles_Dataset.xlsx", "Measles_Case_Records")
annual_trend_2010_2025 <- read_xlsx("Utopia_Measles_Dataset.xlsx", "Annual_Trend_2010_2025")
outbreak_detail <- read_xlsx("Utopia_Measles_Dataset.xlsx", "Outbreak_Detail")
sspe_registry <- read_xlsx("Utopia_Measles_Dataset.xlsx", "SSPE_Registry")
vaccination_coverage <- read_xlsx("Utopia_Measles_Dataset.xlsx", "Vaccination_Coverage")
public_opinion_survey <- read_xlsx("Utopia_Measles_Dataset.xlsx", "Public_Opinion_Survey")
county_geographic_summary <- read_xlsx("Utopia_Measles_Dataset.xlsx", "County_Geographic_Summary")
given_summary_statistics <- read_xlsx("Utopia_Measles_Dataset.xlsx", "Summary_Statistics")
```

## general visualizations

```{r}
annual_trend_2010_2025<- annual_trend_2010_2025 |> 
  mutate(Outbreak_Case_percentage = Outbreak_Linked_Cases / Total_Cases * 100)
```
#Total Cases line
```{r}
annual_trend_2010_2025 |> 
  ggplot(aes(x = Year, y = Total_Cases, label = Total_Cases))+
  geom_line(color = "black", linewidth = 2, lineend = "round")+
  geom_label(fill = "white", color = "blue",
             nudge_y = 5)+
  geom_point(size = 4)+
  scale_x_continuous(breaks = seq(min(annual_trend_2010_2025$Year), max(annual_trend_2010_2025$Year), by = 1),
                     minor_breaks = NULL)+
  scale_y_continuous(breaks = seq(0, max(90), by = 10),
                     minor_breaks = NULL)+
  labs(title = "Total Cases Per Year in Utopia",
       y = "Total Cases")

  #ggsave("total_cases_line.png")
```

#Total cases w/ outbreaks highlighted
```{r}
annual_trend_2010_2025 |> 
  ggplot(aes(x = Year, y = Total_Cases, color = Outbreak_Year, label = Outbreak_Linked_Cases))+
  geom_line(color = "black", linewidth = 2, lineend = "round")+
  geom_label(aes(label = ifelse(Outbreak_Linked_Cases == 0 & Outbreak_Year == "No", NA, paste0(Outbreak_Linked_Cases, " Outbreak \nLinked Cases"))),
             nudge_y = 6)+
  geom_point(size = 4)+
  scale_color_manual(values = c("Yes" = "darkorange3"))+
  scale_x_continuous(breaks = seq(min(annual_trend_2010_2025$Year), max(annual_trend_2010_2025$Year), by = 1),
                     minor_breaks = NULL)+
    scale_y_continuous(breaks = seq(0, max(90), by = 10),
                     minor_breaks = NULL)+
  labs(title = "Outbreak Linked Cases Per Year in Utopia",
       y = "Total Cases")+
  theme(legend.position = "none")

  #ggsave("outbreak_linked_cases_line.png")
```


## Vaccine opinion by county
```{r}
county_geographic_summary <- clean_names(county_geographic_summary)
county_geographic_summary

```
```{r}
county_geographic_summary |> 
  mutate(pct_hospitalized = hospitalizations / total_cases_2010_2025) |> 
  mutate(hospitalizations_per_100k = hospitalizations / estimated_population * 100000)
```

# percent unvax vs complications by county
```{r}
county_geographic_summary |>
  pivot_longer(cols = c(pct_with_complications, pct_unvaccinated_cases),
               names_to = "metric",
               values_to = "Percentage") |>
  ggplot(aes(x = reorder(county, Percentage, decreasing = F), y = Percentage, fill = metric)) +
  geom_col(position = "dodge")+
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), position = position_dodge(width = 0.9), xjust = 2)+
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, by = 10))+
  scale_fill_manual(values = c("orchid3", "olivedrab3"))+
  coord_flip()+
  labs(x = "County", y = "Percentage of Cases")
```


```{r}
county_geographic_summary |> 
  select(county, pct_pro_vaccine_attitude, pct_vaccine_hesitant) |> 
  pivot_longer(cols = c(pct_vaccine_hesitant, pct_pro_vaccine_attitude),
               names_to = "attitude",
               values_to = "percent") |> 
  ggplot(aes(x = reorder(county, percent, decreasing = T), y = percent, fill = attitude))+
  geom_col(position = "dodge") +
  scale_fill_discrete(labels = c("pct_vaccine_hesitant" = "Vaccine Hesitant",
                                 "pct_pro_vaccine_attitude" = "Pro Vaccine"))+
  coord_flip() +
  labs(y = "Percent of Population", 
       x = "County", 
       fill = "Vaccine Attitude",
       title = "Vaccine Attitudes by County") 

## Need to add anti-vax as a category maybe
```
## Vaccine coverage disparities

```{r}
unvax_individual_demographics<- vaccination_coverage |> 
  select(County, Urban_Rural, SES_Category, Insurance_Type, Age_Years, Race_Ethnicity, MMR_Vaccination_Status, Vaccine_Exemption_Type, Reason_if_Unvaccinated, Provider_Recommended_MMR, Received_MMR_at) |> 
  filter(!is.na(Reason_if_Unvaccinated))
```
```{r}
unvax_individual_demographics
#vacc coverage has like 2000 entries but total cases is like 308
```

```{r}
unvax_individual_demographics |> 
  count(Reason_if_Unvaccinated, sort = T) |> 
  mutate(percentage = n / sum(n) *100)
```

```{r}
unvax_individual_demographics |> 
  filter(Reason_if_Unvaccinated == "Cost") |> 
  count(Race_Ethnicity, sort = T) |> 
  mutate(percentage = n / sum(n) *100)
```

```{r}
measles_case_records |> 
  count(Race_Ethnicity, sort = T) |> 
  mutate(percentage = n / sum(n) *100)
```


# kinda unclear the number of ppl in each race just given percents
#total pop of utopia: 310000 (from excel)
#white non hisp = 75.1% 
#black = 1.4%
# native = 5.7%
# asian = 1.4%
# pacific islander = 0.4%
# multiracial = 4.6%
#hispanic = 15.8%
# adds up to 104.4 % though?


```{r}
measles_case_records |> 
  count(Insurance_Type) |> 
  mutate(percentage = n / sum(n) *100) |> 
  arrange(desc(percentage))
```

```{r}
measles_case_records |> 
  count(SES_Category, sort = T) |> 
  mutate(percentage = n / sum(n) *100)
```
```{r}
measles_case_records |> 
  group_by(County, SES_Category) |> 
  summarise(n = n(), .groups = "drop") |> 
  group_by(County) |> 
  mutate(pct = n / sum(n) * 100)

```

```{r}
measles_case_records |> 
  count(Vaccination_Status, sort = T) |> 
  mutate(percentage = n / sum(n) *100)
```
```{r}
vaccination_coverage |> 
  filter(Age_Years <= 2.92) |> 
  count(MMR_Vaccination_Status, sort = T) |> 
  mutate(percentage = n / sum(n) *100)
```
# Public opinion vaccine categories

```{r}
#In presentation state the cut offs for each vaccine attitude category
anti_vax_opinion_survey<- public_opinion_survey |> 
  filter(Vaccine_Attitude_Category == "Anti-vaccine")
  
```
```{r}
vax_hesitant_opinion_survey<- public_opinion_survey |> 
  filter(Vaccine_Attitude_Category == "Vaccine-hesitant")
```
```{r}
pro_vax_opinion_survey<- public_opinion_survey |> 
  filter(Vaccine_Attitude_Category == "Pro-vaccine")
```

#Anti-vax
```{r}
anti_vax_opinion_survey |> 
  count(Education_Level, sort = T) 
```
```{r}
anti_vax_opinion_survey |> 
  count(Info_Source_Primary, sort = T) |> 
  mutate(percentage = n / sum(n) *100)
```

```{r}
anti_vax_opinion_survey |> 
  count(Believes_MMR_Causes_Autism, sort = T) |> 
  mutate(percentage = n / sum(n) *100)
```
```{r}
anti_vax_opinion_survey |> 
  summarise(across(where(is.numeric), mean, na.rm = TRUE))
```
```{r}
anti_vax_opinion_survey |> 
  count(Children_Up_to_Date_on_Vaccines, sort = T) |> 
  mutate(percentage = n / sum(n) *100)
```

#Vax-hesitant

```{r}
vax_hesitant_opinion_survey |> 
  summarise(across(where(is.numeric), mean, na.rm = TRUE))
```
```{r}
vax_hesitant_opinion_survey |> 
  count(Education_Level, sort = T) 
```
```{r}
vax_hesitant_opinion_survey |> 
  count(Info_Source_Primary, sort = T)
```
```{r}
vax_hesitant_opinion_survey |> 
  count(Believes_MMR_Causes_Autism, sort = T) |> 
  mutate(percentage = n / sum(n) *100)
```
```{r}
vax_hesitant_opinion_survey |> 
  count(Children_Up_to_Date_on_Vaccines, sort = T) |> 
  mutate(percentage = n / sum(n) *100)
```
# Pro-vax

```{r}
pro_vax_opinion_survey |> 
  summarise(across(where(is.numeric), mean, na.rm = TRUE))
```
```{r}
pro_vax_opinion_survey |> 
  count(Education_Level, sort = T) 
```
```{r}
pro_vax_opinion_survey |> 
  count(Info_Source_Primary, sort = T)
```
```{r}
pro_vax_opinion_survey |> 
  count(Believes_MMR_Causes_Autism, sort = T) |> 
  mutate(percentage = n / sum(n) *100)
```
```{r}
pro_vax_opinion_survey |> 
  count(Children_Up_to_Date_on_Vaccines, sort = T) |> 
  mutate(percentage = n / sum(n) *100)
```